---
title: "CUNY SPS DATA 621 - CTG5 - HW2"
author: "Betsy Rosalen, Gabrielle Bartomeo, Jeremy O'Brien, Lidiia Tronina, Rose Koh"
date: "March 13, 2019"
output:
    pdf_document:
        toc: true
        toc_depth: 2
        number_sections: true
        fig_width: 8
        fig_height: 8
        fig_caption: true
        highlight: haddock
        df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, echo=FALSE, message=FALSE, warning=FALSE)
chooseCRANmirror(graphics=FALSE, ind=1)
if (!require('data.table')) (install.packages('data.table'))
if (!require('ggplot2')) (install.packages('ggplot2'))
if (!require('caret')) (install.packages('caret'))
if (!require('pROC')) (install.packages('pROC'))
```

Deliverables should use R functions and the other packages to generate the `classification metrics` for the provided data set.


__1. Download the classification output data set (attached in Blackboard to the assignment).__

```{r include=FALSE}
# load data
raw.data <- read.csv('https://raw.githubusercontent.com/silverrainb/data621proj2/master/classification-output-data.csv', stringsAsFactors = F, header = T)
data <- as.data.table(raw.data)
```

__2. The data set has three key columns we will use:__

* class: the actual class for the observation
* scored.class: the predicted class for the observation (based on a threshold of 0.5)
* scored.probability: the predicted probability of success for the observation

__Use the table() function to get the raw confusion matrix for this scored dataset. Make sure you understand the output. In particular, do the rows represent the actual or predicted class? The columns?__

```{r}
tbl <- table(data$scored.class, 
                   data$class) 
knitr::kable(tbl)
```

* rows = predicted, cols = actual
* 1 is positive, 0 is negative

```{r}
# 3. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the accuracy of the predictions.
# 
# 4. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the classification error rate of the predictions.
# 
# Verify that you get an accuracy and an error rate that sums to one.
# 
# 5. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the precision of the predictions.
# 
# 6. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the sensitivity of the predictions. Sensitivity is also known as recall.
# 
# 7. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the specificity of the predictions.
# 
# 8. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the F1 score of the predictions.
```
__3.-8.: Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns (3)Accuracy; (4)Error rate; (5)Precision; (6)Sensitivity(recall); (7)Specificity; (8)F1 score of the predictions. Verify that you get an accuracy and an error rate that sums to one.__
 
```{r}
# Done by Rose, modified slightly by Gabby to be in line with the instructions (takes the data set as a dataframe)
ClassCalc <- function(df_for_calcs) {
  if (!is.data.frame(df_for_calcs)) stop("The function ClassCalc requires a dataframe of two columns. The first should be the predicted values, and the second the actual values.") # Gabby added this
  tbl <- table(df_for_calcs[,1], df_for_calcs[,2])
  # with tbl, get TP, FP, FN, TN
  tp <- tbl[1,1];
  fp <- tbl[1,2];
  fn <- tbl[2,1];
  tn <- tbl[2,2];
  
  # calculate summary stats
  accuracy <- (tp+tn) / (tp+fp+tn+fn)
  error.rate <- (fp+fn) / (tp+fp+tn+fn)
  precision <- tp / (tp+fp)
  sensitivity <- tp / (tp+fn)
  specificity <- tn / (tn+fp)
  f1 <- (2 * precision * sensitivity) / (precision + sensitivity)
  
  # create dataframe
  df <- data.frame(accuracy = accuracy,
                   error.rate = error.rate,
                   precision = precision,
                   sensitivity = sensitivity,
                   specificity = specificity,
                   f1 = f1)
  return(df)
}

output <- ClassCalc(raw.data[,c("scored.class", "class")])

print(output$accuracy + output$error.rate)

knitr::kable(output)
```


__9. Before we move on, let's consider a question that was asked: What are the bounds on the F1 score? Show that the F1 score will always be between 0 and 1.(Hint: If 0 < a < 1 and 0 < b < 1 then ab < a)__

```{r}
# GABBY GOT THIS ONE!
```

Given the formula for the F1 score...

$$\text{F1 Score} = \frac{2 \times \text{Precision} \times \text{Sensitivity}}{\text{Precision} + \text{Sensitivity}}$$

... it can be determined the values of Precision and Sensitivity are of utmost importance.

$$\text{Precision} = \frac{\text{TP}}{\text{TP}+\text{FP}} \text{   ,  } \text{Sensitivity} = \frac{\text{TP}}{\text{TP}+\text{FN}}$$

Considering the extremes can lead to a more whole answer regarding the bounds on the F1 score. First, if we assume TP has a value of 0...

$$\text{Precision} = \frac{0}{0 +\text{FP}} \text{   ,  } \text{Sensitivity} = \frac{0}{0+\text{FN}}$$

$$\text{Precision} = \frac{0}{\text{FP}} \text{   ,  } \text{Sensitivity} = \frac{0}{\text{FN}}$$

$$\text{Precision} = 0 \text{   ,  } \text{Sensitivity} = 0$$

Next, if it is assumed TP has a value of 1...

$$\text{Precision} = \frac{1}{1 +\text{FP}} \text{   ,  } \text{Sensitivity} = \frac{1}{1+\text{FN}}$$

Depending on the values of FP and FN for Precision and Sensitivity respectively, this will always result in a value greater than 0 and less than or equal to 1. This is the case for all non-zero values of TP.

Since it has been proven that Precision and Sensitivity will have values between 0 and 1 inclusively, the same treatment may be applied to the F1 Score.

If Precision and Sensitivity are both 0...

$$\text{F1 Score} = \frac{2 \times \text{Precision} \times \text{Sensitivity}}{\text{Precision} + \text{Sensitivity}}$$

$$\text{F1 Score} = \frac{2 \times 0 \times 0}{0 + 0}$$

$$\text{F1 Score} = \frac{0}{0}$$

$$\text{F1 Score} = 0$$

If Precision and Sensitivity are both 1...

$$\text{F1 Score} = \frac{2 \times \text{Precision} \times \text{Sensitivity}}{\text{Precision} + \text{Sensitivity}}$$

$$\text{F1 Score} = \frac{2 \times 1 \times 1}{1 + 1}$$

$$\text{F1 Score} = \frac{2}{2}$$

$$\text{F1 Score} = 1$$

On each extreme the F1 Score has a value of 0 and a value of 1. That means if Precision and Sensitivity should fall between 0 and 1, the F1 Score should also have a value that falls between 0 and 1. 

_In short, the bounds of the F1 score are 0 and 1._

__10. Write a function that generates an ROC curve from a data set with a true classification column (class in our example) and a probability column (scored.probability in our example). Your function should return a list that includes the plot of the ROC curve and a vector that contains the calculated area under the curve (AUC). Note that I recommend using a sequence of thresholds ranging from 0 to 1 at 0.01 intervals.__

__11. Use your <b>created R functions</b> and the provided classification output data set to produce all of the classification metrics discussed above.__

__12. Investigate the <b>caret</b> package. In particular, consider the functions confusionMatrix, sensitivity, and specificity. Apply the functions to the data set. How do the results compare with your own functions?__

__13. Investigate the <b>pROC</b> package. Use it to generate an ROC curve for the data set. How do the results compare with your own functions?__